---
title: "R Notebook"
output: Box Crab
---

Goal: pull data about flamed box crab and compile it to one data table.

```{r setup}
load("glf_reduced.RData")
load("seus_reduced.RData")
load("neus_reduced.RData")
load("GSDET_CAT_INF_SPE_merge.RData")

library(dplyr)
library(data.table)
install.packages("stringr")
library(stringr)
install.packages("tibble")
library(tibble)

capitalize <- function(x){
first <- toupper(substr(x, start=1, stop=1)) ## capitalize first letter
rest <- tolower(substr(x, start=2, stop=nchar(x)))## everything else lowercase
paste0(first, rest)
}
```

Make spelling edits to Gulf data
```{r spelling}
glf_reduced.dt$SPEC_GLF <- str_replace_all(glf_reduced.dt$SPEC_GLF, "FLAMME", "FLAMMEA")
```

Extract relavent data.
```{r extraction}
glf_boxcrab <- subset(glf_reduced.dt, SPEC_GLF == "FLAMMEA")
seus_boxcrab <- subset(seus_reduced, SPECIESSCIENTIFICNAME.x == "CALAPPA FLAMMEA")
neus_boxcrab <- subset(neus_reduced, SCINAME == "CALAPPA FLAMMEA")
```

Reduce datasets to have same data.
```{r reduce}
glf_boxcrab_red <- glf_boxcrab[, c("LEN_GLF", "SEX_GLF", "MO_DAY_YR", "TEMP_BOT", "DECSLAT", "DECSLON")]
seus_boxcrab <- as.data.table(seus_boxcrab)
seus_boxcrab_red <- seus_boxcrab[, c("DATE.x", "SPECIESSCIENTIFICNAME.x", "SPECIESCOMMONNAME.x", "TEMPBOTTOM.x", "LATITUDESTART.x", "LONGITUDESTART", "LEN1", "SEXDESCRIPTION")]
neus_boxcrab_red <- neus_boxcrab[, c("CATCHSEX", "LAT", "LON", "EST_TOWDATE", "BOTTEMP", "LENGTH", "SCINAME", "COMNAME")]
```

Add and rename columns so they can bind.
```{r add and rename}
glf_boxcrab_red <- add_column(glf_boxcrab_red, SCINAME = "Calappa flammea")
glf_boxcrab_red <- add_column(glf_boxcrab_red, COMNAME = "Flame box crab")
glf_boxcrab_red <- rename(glf_boxcrab_red, "LENGTH" = "LEN_GLF", "SEX" = "SEX_GLF", "TEMPERATURE" = "TEMP_BOT", "LATITUDE" = "DECSLAT", "LONGITUDE" = "DECSLON", "DATE" = "MO_DAY_YR")
seus_boxcrab_red <- rename(seus_boxcrab_red, "DATE" = "DATE.x", "SCINAME" = "SPECIESSCIENTIFICNAME.x", "COMNAME" = "SPECIESCOMMONNAME.x", "TEMPERATURE" = "TEMPBOTTOM.x", "LATITUDE" = "LATITUDESTART.x", "LONGITUDE" = "LONGITUDESTART", "LENGTH" = "LEN1", "SEX" = "SEXDESCRIPTION")
seus_boxcrab_red$SCINAME <- capitalize(seus_boxcrab_red$SCINAME)
seus_boxcrab_red$COMNAME <- capitalize(seus_boxcrab_red$COMNAME)
neus_boxcrab_red <- rename(neus_boxcrab_red, "SEX" = "CATCHSEX", "LATITUDE" = "LAT", "LONGITUDE" = "LON", "DATE" = "EST_TOWDATE", "TEMPERATURE" = "BOTTEMP")
neus_boxcrab_red$SCINAME <- as.character(neus_boxcrab_red$SCINAME)
neus_boxcrab_red$SCINAME <- capitalize(neus_boxcrab_red$SCINAME)
neus_boxcrab_red$COMNAME <- as.character(neus_boxcrab_red$COMNAME)
neus_boxcrab_red$COMNAME <- capitalize(neus_boxcrab_red$COMNAME)
```

Reorder columns so they can bind.
```{r reorder}
glf_boxcrab_red <- glf_boxcrab_red[,c("SCINAME", "COMNAME", "LENGTH", "SEX", "LATITUDE", "LONGITUDE", "TEMPERATURE", "DATE")]
seus_boxcrab_red <- seus_boxcrab_red[,c("SCINAME", "COMNAME", "LENGTH", "SEX", "LATITUDE", "LONGITUDE", "TEMPERATURE", "DATE")]
neus_boxcrab_red <- neus_boxcrab_red[,c("SCINAME", "COMNAME", "LENGTH", "SEX", "LATITUDE", "LONGITUDE", "TEMPERATURE", "DATE")]

##FOR EACH COMMAND HERE ZJK ADDED A COMMA BEFORE c("SCINAME"...etc) to specify that we want to keep these columns (data table syntax = [row,column])


```

Bind tables.
```{r bind}

##I hashed out the 3 commands below because these objects should already be data tables
#glf_boxcrab_red.dt <- as.data.table(glf_boxcrab_red)
#seus_boxcrab_red.dt <- as.data.table(seus_boxcrab_red)
#neus_boxcrab_red.dt <- as.data.table(neus_boxcrab_red)
boxcrab_table <- rbind(glf_boxcrab_red, neus_boxcrab_red, seus_boxcrab_red)
#I couldn't get this to work either, see below


class(glf_boxcrab_red)
```
Ria: I tried to perform the function above, and got an error message (probably the same you were getting), I think it's because the class of each of the columns is different depending on the region. For example,
```{r structure for neus}
str(neus_boxcrab_red)
```

'data.frame':	4 obs. of  8 variables:
 $ SCINAME    : chr  "Calappa flammea" "Calappa flammea" "Calappa flammea" "Calappa flammea"
 $ COMNAME    : chr  "Flame box crab" "Flame box crab" "Flame box crab" "Flame box crab"
 $ LENGTH     : num  NA 2 8 10
 $ SEX        : int  0 0 0 0
 $ LATITUDE   : num  36.4 36 39.8 33.3
 $ LONGITUDE  : num  -75.7 -75.3 -72 -77.7
 $ TEMPERATURE: num  9.5 6.8 8.3 14
 $ DATE       : POSIXct, format: "1976-03-21 22:56:00" "1977-03-23 13:29:00" "1977-04-04 13:57:00" "1981-03-24 23:09:00"
 
But for SEUS, 
```{r structure for seus}
str(seus_boxcrab_red)
```
Classes ‘data.table’ and 'data.frame':	13096 obs. of  8 variables:
 $ SCINAME    : chr  "Calappa flammea" "Calappa flammea" "Calappa flammea" "Calappa flammea" ...
 $ COMNAME    : chr  "Flame box crab" "Flame box crab" "Flame box crab" "Flame box crab" ...
 $ LENGTH     : chr  "146" "131" "134" "136" ...
 $ SEX        : chr  "FEMALE" "MALE" "MALE" "MALE" ...
 $ LATITUDE   : chr  "32.4780" "32.4780" "32.4780" "32.4780" ...
 $ LONGITUDE  : chr  "-80.1380" "-80.1380" "-80.1380" "-80.1380" ...
 $ TEMPERATURE: chr  "14.70" "14.70" "14.70" "14.70" ...
 $ DATE       : chr  "04-13-1989" "04-13-1989" "04-13-1989" "04-13-1989" ...
 - attr(*, ".internal.selfref")=<externalptr> 
 
 **Before you try to rbind the rows, they structure for all three should be (I've written out the full names, but you can see the abbreviations to expect above):**
 Classes ‘data.table’ and 'data.frame':	13096 obs. of  8 variables:
 $ SCINAME    : character  
 $ COMNAME    : character  
 $ LENGTH     : numeric  
 $ SEX        : factor  
 $ LATITUDE   : numeric  
 $ LONGITUDE  : numeric  
 $ TEMPERATURE: numeric  
 $ DATE       : date  
 
If you play around with as.numeric(), as.factor(), and as.Date() functions on columns before binding, this should solve your problem. Converting the date may be a little tricky, but this [website](https://www.statmethods.net/input/dates.html) may be helpful.

 